\ourtitle

### 2020-05-19 paper/article.tex
Some adjustments to the paper

\section{Introduction}
Musikla stands for Music and Keyboard Language. Our goal is to develop a DSL (\textit{Domain Specific Language}) that allows treating musical events as regular data in a programming language. More than generating these musical events offline, we want to be able to easily declare keyboards that map keys and expressions to mutate the state or play musical events (or both).

The project can be sectioned in three different, modular layers: inputs, the language, and outputs. While music events can be described as code literals in our language, they can also originate from many other sources. After being processed by our language, it then emits a stream of musical events to the \textbf{Player}, which then multiplexes those events into however many outputs the user defined.

@@

While the development of both the input and output layers, as well as their many respective components presents by itself many interesting challenges that could be discussed, we will instead focus this paper on the aspects of the middle layer: the \textit{interpreter}, while ackowledging the existence (and their effects) of the layers that wrap around it.

As such, the problem of developing the interpreter can be divided into two parts: the syntax used for describing the notes and the operators the compose them inside the language; and the semantics of the generated events, how they are stored in memory, and how their temporal properties are handled without forcing the programmer/user to manually type them.

Designing the syntax for describing those musical expressions, especially given our strong desire to make those musical expressions first class citizens, like other primitive data types in most programming languages (such as numbers, strings or arrays are), did unearth some challenges. To minimize the learning curve for new users, and avoid reinventing the wheel, we decided to adopt a subset of the very popular note declaration syntax from the ABC Notation project\cite{AbcNotation} and integrate it with our language.

@@

The simplest way of generating such musical events in a programming language is to use already common, \textit{low-level}, programming mechanisms, such as using a procedural approach where the user creates each event manually by calling a function and providing as parameters all the events' information, such as it's timestamp and duration. This is the approach used by some of the existing languages in this space, such as \textit{SonicPi}.

@@

A musical \textbf{Event} can be one of many things, such as a \textit{note}, a \textit{chord}, or even more implementation-specific events like MIDI messages. While all events must have a start time, some events can be instantaneous (events with a duration of zero time units).

%\begin{itemize}
% \item \textbf{NoteEvent} Describes a single note
% \item \textbf{ChordEvent} Describes a chord (group of notes to play in parallel). Is equivalent to having multiple separate \textbf{NoteEvent}'s with the same timestamp, but conveys a more specific semantic value.
% \item \textbf{ControlChangeEvent}
% \end{itemize}

@@

One of the features that Python boasts (but are certainl not exclusive to it) that have eased our implementation are generators\cite{PEP255}. They integrate very nicely into both our concept of emitting musical events as sequences (or iterators, as they are called in Python), as well as into our concept of laziness, where events are generated on demand when needed, and thus infinite musical sequences can be handled easily.

\paragraph*{Context State}
To keep track of the \textit{cursor} (the current timestamp where the next event should start) each operator in our language is implemented as a function call that receives an implicit \texttt{Context} object. While here we'll mostly focus just on the methods related to time management provided by the context, it can be used to store other types of information, like the default length of a musical note, for instance, to avoid forcing the user to type it out all the time.

@@

\paragraph*{Concatenation}
We've seen how single events' creation is handled. Now it is important for us to see how we can combine those events together. And probably the most straightforward operator of all, concatenation, it simply consumes each event. Each event, as we've seen before, is responsible for seeking the context depending on the event's duration.
\begin{lstlisting}[caption={Snippet of Wet Hands by C418},label=list:4,captionpos=t,abovecaptionskip=-\medskipamount]
S4/4 T74 L/8 V90;
A, E A B ^c B A E D ^F ^c e ^c A3;
\end{lstlisting}

\begin{figure}[ht]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/concatenation2.png}}%
  }%
  \caption{Generated music sheet for concatenation\protect\footnotemark.}
  \label{fig:concatenation}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/repetition2.png}}%
  }%
  \caption{Generated music sheet for repetition.}
  \label{fig:repetition}
\end{figure}

@@

\begin{lstlisting}[caption={Snippet of Soft to Be Strong by Marina},label=list:6,captionpos=t,abovecaptionskip=-\medskipamount]
T120 V70 L1;
r/4 ^g/4 ^g/4 ^g/4   ^f/2 e/8 ^d3/8   ^c2 | [^Cm] [BM]  [AM] [BM] 
\end{lstlisting}

\begin{figure}[ht]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/parallel2.png}}%
  }%
  \caption{Generated music sheet for parallel}
  \label{fig:parallel}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/fugue.png}}%
  }%
  \caption{Generated music sheet for fugue example}
  \label{fig:fugue}
\end{figure}
% https://drive.google.com/open?id=1-igVwdhXMxOSE1cMg_8HimDOXMhsY9US

\paragraph*{Grids}
To define a grid there is only one parameter required: the length of it's cells. When aligning musical events, anything inside each cell will be pushed to the closest edge of the cell.

>>>>

\section{Introduction}
Musikla stands for Music and Keyboard Language. Our goal is to develop a DSL (\textit{Domain Specific Language}) that allows treating musical events as regular data in a programming language. More than generating these musical events offline, we want to be able to easily declare keyboards that map keys to expressions that either mutate the state or play musical events (or even both).

The project can be partitioned in three different, modular layers: inputs, the language, and outputs. While music events can be described as code literals inside our language, they can also originate from many other sources (such as files or physical devices such as pianos). After being processed by our language, they are then emitted as a stream of musical events to the \textbf{Player} component, which then multiplexes those events into however many outputs the user defined.

@@

While the development of both the input and output layers, as well as their many respective components, presents by itself many interesting challenges that could be discussed, we will instead focus this paper on the aspects of the middle layer: the \textit{interpreter}, while ackowledging the existence (and their effects) of the layers that wrap around it.

As such, the problem of developing the interpreter can be divided into two parts: the syntax used for describing the notes and the operators that compose them inside the language; and the semantics of the generated events, how they are stored in memory, and how their temporal properties (start time and duration) are handled without forcing the programmer/user to always manually type them.

Designing the syntax for describing those musical expressions, especially given our strong desire to make those musical expressions first class citizens like other primitive data types in most programming languages (such as numbers, strings or arrays are), did unearth some challenges. To minimize the learning curve for new users, and avoid reinventing the wheel, we decided to adopt a subset of the very popular note declaration syntax from the ABC Notation project\cite{AbcNotation} and integrate it with our language.

@@

The simplest way of generating such musical events in a programming language is to use already common, \textit{low-level}, programming mechanisms, such as using a procedural approach where the user creates each event manually by calling a function and providing as parameters all the events' information, such as it's timestamp and duration. This is the approach used by some of the existing languages in this space, such as \textit{SonicPi}\cite{Aaron2014TemporalSF}.

@@

A musical \textbf{Event} can be one of many things, such as a \textit{note}, a \textit{chord}, or even more implementation-specific events like MIDI messages\cite{Loy1985MusiciansMA}. While all events must have a start time, some events can be instantaneous (events with a duration of zero time units).

@@

One of the features that Python boasts (but are certainly not exclusive to it) that have eased our implementation of the language are generators\cite{PEP255}. They integrate very nicely into both our concept of emitting musical events as sequences (or iterators, as they are called in Python and other languages), as well as into our concept of laziness, where events are generated on demand when needed, and thus infinite musical sequences can be handled easily.

\paragraph*{Context State}
To keep track of the \textit{cursor} (the current timestamp where the next event should start) each operator in our language is implemented as a function call that receives an implicit \texttt{Context} object. While here we'll mostly focus just on the methods related to time management provided by the context, it can be used to store other types of information, like the default length of a musical note, for instance, to avoid forcing the user to type it out all the time, or the tempo at which it is to be played.

@@

\paragraph*{Concatenation}
We've seen how single events' creation is handled. Now it is important for us to see how we can combine those events together. And probably the most straightforward operator of all, concatenation, it simply consumes each event. Each event, as we've seen before, is responsible for seeking the context depending on the event's duration.
\begin{lstlisting}[caption={Snippet of the song \textit{Wet Hands} by C418},label=list:4,captionpos=t,abovecaptionskip=-\medskipamount]
S4/4 T74 L/8 V90;
A, E A B ^c B A E D ^F ^c e ^c A3;
\end{lstlisting}

\begin{figure}[ht]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/concatenation2.png}}%
  }%
  \caption{Generated music sheet for concatenation\protect\footnotemark, audio version available \href{https://drive.google.com/open?id=1TP4lcul81s8iMCUFmD3HKnSpeftCzKT0}{\underline{here}}.}
  \label{fig:concatenation}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/repetition2.png}}%
  }%
  \caption{Generated music sheet for repetition, audio version available \href{https://drive.google.com/open?id=1IIm8PQkLsNFMK9MNSVubJSG6SP6KwhPL}{\underline{here}}.}
  \label{fig:repetition}
\end{figure}

@@

\begin{lstlisting}[caption={Snippet of the song \textit{Soft to Be Strong} by Marina},label=list:6,captionpos=t,abovecaptionskip=-\medskipamount]
T120 V70 L1;
r/4 ^g/4 ^g/4 ^g/4   ^f/2 e/8 ^d3/8   ^c2 | [^Cm] [BM]  [AM] [BM] 
\end{lstlisting}

\begin{figure}[ht]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/parallel2.png}}%
  }%
  \caption{Generated music sheet for parallel, audio version available \href{https://drive.google.com/open?id=1ENTm3hZonYHyQIOgRZ8TQ1Qz-AfRLt2I}{\underline{here}}.}
  \label{fig:parallel}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  {%
  \setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0pt}%
  \fbox{\includegraphics[width=.85\linewidth]{../code/musikla/musikla/examples/paper/fugue.png}}%
  }%
  \caption{Generated music sheet for fugue example, audio version available \href{https://drive.google.com/open?id=1dIfvnhhKn73Vpp0W6ss6RLsv6PQ_HFTF}{\underline{here}}.}
  \label{fig:fugue}
\end{figure}
% https://drive.google.com/open?id=1-igVwdhXMxOSE1cMg_8HimDOXMhsY9US

\paragraph*{Grids}
To define a grid there is only one parameter required: the length of it's cells. When aligning musical events, anything that falls inside each cell will be pushed to the closest edge of the cell.
